\documentclass[10pt,conference]{IEEEtran} 
\usepackage{multicol} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath} 
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{url}

\begin{document}

\title{\vfill NLCBSMM: A Virtual Memory Manager in Userspace} 
\author{Ryan Verdon, Stephen Holsapple}
\date{\today}
\maketitle

\def \projname {NLCBSMM}
\newtheorem{definition}{Definition}[section]
\newtheorem{condition}{Condition}[section]

\begin{abstract}
This paper presents \projname{}: a lazy release consistent distributed shared memory (DSM) system built entirely in user-space.  The \projname{} system is capable of executing threaded applications implemented with \verb,pthreads, on a cluster of networked machines without \em any \em modifications to the target application, making the system a general DSM solution compared to others that require specially written applications or memory annotations.  The \projname{} system features a centralized memory manager~\cite{Li:1989:MCS:75104.75105} built atop Hoard~\cite{Berger:1999:HFS:899944, Berger:2000:HSM:356989.357000}: a fast, scalable, and memory-efficient allocator for shared-memory multiprocessors.

In our analysis, we discuss various memory coherence models and their applicability to existing software distributed shared memory (SDSM) systems, we discuss how SDSM systems can be straight-forward to implement in a Linux environment, and lastly, we present preliminary performance characteristics that the release consistent distributed shared memory system exhibits.  In our presentation of performance characteristics, we show that network faults can be resolved in as little as 3 milliseconds with our implementation, we provide a comparison between runtime performance between \projname{} and the system provided non-networked version of \verb,malloc,, and lastly a comparison between various memory coherency models.

\end{abstract}

\section{Introduction}

Distributed shared memory (DSM) systems allow many physically separate processes to see identical address spaces.  This allows the processes to use shared memory techniques to pass messages through memory rather than sending messages to specific processes.  DSM is also good for processing large complex data sets without having to replicate the entire data set on every node processing data.  Only the data that is currently being worked on has to be present physically on the machine.  We decided to implement a software distributed shared memory (SDSM) in user-space because this allowed us to not have to modify the OS kernel.  This way \projname{} can be used on a large variety of systems.  

The rest of the paper is organized as follows.  Section 2 provides some background and discusses related work to our project.  Section 3 provides a discussion of memory consistency models and most importantly presents the idea of release consistency.  Section 4 presents the design of our project along with the assumptions that were made.  Section 5 describes the implementation of \projname{}.  Finally, section 6 presents a comparison of \projname{} to \verb,malloc, along with results on how long network faults take.

\input{nlcbsmm-background.tex}
\input{nlcbsmm-design.tex}
\input{nlcbsmm-impl.tex}
\input{nlcbsmm-results.tex}


\bibliographystyle{IEEEannot}
\bibliography{nlcbsmm}
\end{document}
